---
title: 'Final Project Analysis : R Scripts'
author: "Antara Saha"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
---

# **INTRODUCTION**:

For my final project, I will be performing analysis to find the relation between Homeschooling with vaccination status, race & education attainment.

For this analysis, I will use the Household Pulse Survey data from Phase 3.4 and Phase 3.5, available here:

* [Phase 3.5 PUF Releases](https://www.census.gov/programs-surveys/household-pulse-survey/datasets.html#phase3.5)
* [Phase 3.4 PUF Releases](https://www.census.gov/programs-surveys/household-pulse-survey/datasets.html#phase3.4)


### **STEP 1: LOAD DATA INTO R**

```{r message=FALSE}
# Load Data into R
library(tidyverse)
library(dplyr)
library(knitr)
library(lmtest)
library(ggpubr)
library(ggthemes)

inp_fl <- list.files(pattern=".csv")

tmp_dt_5_11 <- data.frame(matrix(ncol = 5, nrow = 0))
colnames(tmp_dt_5_11) <- c('RRACE','EEDUC','ENRHMSCHK',
                           'KIDS_5_11Y','KIDDOSESRV_5_11Y')

tmp_dt_12_17 <- data.frame(matrix(ncol = 5, nrow = 0))
colnames(tmp_dt_12_17) <- c('RRACE','EEDUC','ENRHMSCHK',
                            'KIDS_12_17Y','KIDDOSESRV_12_17Y')

#Load file in R
for (i in 1:length(inp_fl)) 
{
  inp_dt <- data.frame(read.csv(inp_fl[i], header = TRUE, stringsAsFactors = FALSE,
                                na.strings = c("-88","-99")))
  
  if(i<=3){
    inp_dt_5_11 <- inp_dt%>%
      select(RRACE,EEDUC,ENRHMSCHK,KIDS_5_11Y,KIDDOSES_5_11Y)%>%
      filter(KIDS_5_11Y==1)%>%
      mutate(HMSCH = ifelse(ENRHMSCHK ==1,1,0))%>%
      rename(KIDDOSESRV_5_11Y=KIDDOSES_5_11Y)
    
    inp_dt_12_17 <- inp_dt%>%
      select(RRACE,EEDUC,ENRHMSCHK,KIDS_12_17Y,KIDDOSES_12_17Y)%>%
      filter(KIDS_12_17Y==1)%>%
      mutate(HMSCH = ifelse(ENRHMSCHK ==1,1,0))%>%
      rename(KIDDOSESRV_12_17Y=KIDDOSES_12_17Y)
  }
  else{
    inp_dt_5_11 <- inp_dt%>%
      select(RRACE,EEDUC,ENRHMSCHK,KIDS_5_11Y,KIDDOSESRV_5_11Y)%>%
      filter(KIDS_5_11Y==1)%>%
      mutate(HMSCH = ifelse(ENRHMSCHK ==1,1,0))
    
    inp_dt_12_17 <- inp_dt%>%
      select(RRACE,EEDUC,ENRHMSCHK,KIDS_12_17Y,KIDDOSESRV_12_17Y)%>%
      filter(KIDS_12_17Y==1)%>%
      mutate(HMSCH = ifelse(ENRHMSCHK ==1,1,0))
    
  }
  
  tmp_dt_5_11 <- bind_rows(tmp_dt_5_11, inp_dt_5_11)
  tmp_dt_12_17 <- bind_rows(tmp_dt_12_17, inp_dt_12_17)
  
  inp_dt <- data.frame(NULL)
}

# Remove records where the Vaccination Status for kids is anything other than Yes or No.

fnl_dt_5_11 <-
  tmp_dt_5_11%>%
  filter(KIDDOSESRV_5_11Y !=3&!is.na(KIDDOSESRV_5_11Y))

fnl_dt_12_17 <-
  tmp_dt_12_17%>%
  filter(KIDDOSESRV_12_17Y !=3&!is.na(KIDDOSESRV_12_17Y))

fnl_dt_5_11$HMSCH[is.na(fnl_dt_5_11$HMSCH)] <- 0
fnl_dt_12_17$HMSCH[is.na(fnl_dt_12_17$HMSCH)] <- 0
```

### **STEP 2: CREATE FACTORS**

```{r message=FALSE}
# Convert Vaccination Status, Race, and Education attainment into factors. 

# RACE
fnl_dt_5_11$RRACE <- 
  factor(fnl_dt_5_11$RRACE,labels = c("White", "Black", "Asian","Other"))

fnl_dt_12_17$RRACE <- 
  factor(fnl_dt_12_17$RRACE,labels = c("White", "Black", "Asian","Other"))

# EDUCATION ATTAINMENT
fnl_dt_5_11$EEDUC <- ifelse(fnl_dt_5_11$EEDUC <=4 ,"Less Educated","Well Educated")
fnl_dt_12_17$EEDUC <- ifelse(fnl_dt_12_17$EEDUC <=4 ,"Less Educated","Well Educated")

fnl_dt_5_11$EEDUC <- as.factor(fnl_dt_5_11$EEDUC)
fnl_dt_12_17$EEDUC <- as.factor(fnl_dt_12_17$EEDUC)

# VACCINATION STATUS
fnl_dt_5_11$KIDDOSESRV_5_11Y <- 
  factor(fnl_dt_5_11$KIDDOSESRV_5_11Y,labels = c("Vaccinated","Not_Vaccinated"))

fnl_dt_12_17$KIDDOSESRV_12_17Y <- 
  factor(fnl_dt_12_17$KIDDOSESRV_12_17Y,labels = c("Vaccinated","Not_Vaccinated"))
```

Final Data Tables after data cleansing for age groups 5-11 and 12-17 years old.

```{r}
kable(head(fnl_dt_5_11),caption = "**Table 1: Final Data Table for Age Group 5-11 Years Old**")
kable(head(fnl_dt_12_17),caption = "**Table 2: Final Data Table for Age Group 12-17 Years Old**")
```

# **FOR KIDS AGE GROUP 5-11 YEARS OLD**
### **STEP 3: DESCRIPTIVE STATISTICS**

Proportion of kids Homeschooling based on the demographic factors.

```{r message=FALSE, fig.width=10, fig.height=5}
# For Kids Age Group : 5_11 Yrs Old

# Plot Percent of Kids Homeschooling Vs Vaccination Status
plot1 <- fnl_dt_5_11%>%
  group_by(KIDDOSESRV_5_11Y,HMSCH)%>%
  summarise(count = n())%>%
  mutate(pct = (count/sum(count))*100)%>%
  filter(HMSCH == 1)%>%
  ggplot(aes(KIDDOSESRV_5_11Y, pct))+
  geom_bar(stat = "identity", position = 'dodge', fill="steelblue")+
  scale_y_continuous(breaks = seq(0,20,5))+
  labs(y = "% of Kids in Homeschooling", x = "Vaccination Status")+
  geom_text(aes(label = formatC(pct, digits = 1, format = "f")),
            colour = "white", size = 4, vjust = 1.5)+
  theme_light()

# Plot Percent of Kids Homeschooling Vs Race
plot2 <- fnl_dt_5_11%>%
  group_by(RRACE,HMSCH)%>%
  summarise(count = n())%>%
  mutate(pct = (count/sum(count))*100)%>%
  filter(HMSCH == 1)%>%
  ggplot(aes(RRACE, pct))+
  geom_bar(stat = "identity", position = 'dodge', fill="steelblue")+
  scale_y_continuous(breaks = seq(0,20,5))+
  labs(y = "% of Kids in Homeschooling", x = "Race")+
  geom_text(aes(label = formatC(pct, digits = 1, format = "f")),
            colour = "white", size = 4, vjust = 1.5)+
  theme_light()

# Plot Percent of Kids Homeschooling Vs Education Attainment
plot3 <- fnl_dt_5_11%>%
  group_by(EEDUC,HMSCH)%>%
  summarise(count = n())%>%
  mutate(pct = (count/sum(count))*100)%>%
  filter(HMSCH == 1)%>%
  ggplot(aes(EEDUC, pct))+
  geom_bar(stat = "identity", position = 'dodge', fill="steelblue")+
  scale_y_continuous(breaks = seq(0,20,5))+
  labs(y = "% of Kids in Homeschooling", x = "Education Attainment")+
  geom_text(aes(label = formatC(pct, digits = 1, format = "f")),
            colour = "white", size = 4, vjust = 1.5)+
  theme_light()

# Combining all plot into one
combined_plot_5_11 <- ggarrange(plot1, plot2, plot3, nrow = 1, ncol = 3,
                                common.legend = TRUE,legend="bottom")
annotate_figure(combined_plot_5_11,
                top = text_grob("Percent of Kids Homeschool by: Kids Vaccination Status,Race,Education Attainment\nAge Group: 5_11Yrs Old\n",
                                color = "black", face = "bold", size = 14))+
  theme_light()
```

### **STEP 4: DISPROPORTIONALITY INDICES**

Determine the disproportional indices for the % of children who are Homeschooling based on the demographic characteristics 

```{r message=FALSE}
# Define Function to calculate Disproportionality Indices
disp <- function(cats,condition,pop) 
  {
  dispnum <- condition/sum(condition)
  dispdenom <- pop/sum(pop)
  disp <- dispnum/dispdenom
  return(data.frame(category=cats, share_condition=dispnum, 
                    share_population=dispdenom, disproportionality=disp))
}
```
******
```{r message=FALSE}

# Disproportionality Indices for kids Homeschooling of age 5-11 yrs old by Vaccination Status
fnl_dt_5_11_summary <- fnl_dt_5_11%>%
  group_by(KIDDOSESRV_5_11Y)%>%
  summarise(hh_kids = n(),
            hh_kids_homeschool = length(HMSCH[HMSCH == 1]),
            hh_kids_not_homeschool = length(HMSCH[HMSCH == 0]))

disp_ind_5_11 <- disp(fnl_dt_5_11_summary$KIDDOSESRV_5_11Y,fnl_dt_5_11_summary$hh_kids_homeschool,fnl_dt_5_11_summary$hh_kids)

# Rank them from highest to lowest
disp_ind_5_11$Rank <- rank(-disp_ind_5_11$disproportionality)

kable(disp_ind_5_11[order(disp_ind_5_11$Rank, decreasing = FALSE), ],
      caption = "**Table 3: Disproportionality Indices for kids Homeschooling of age 5-11 yrs old by Vaccination Status**")
```
******
```{r message=FALSE}
# Disproportionality Indices for kids Homeschooling of age 5-11 yrs old by Race
fnl_dt_5_11_summary <- fnl_dt_5_11%>%
  group_by(RRACE)%>%
  summarise(hh_kids = n(),
            hh_kids_homeschool = length(HMSCH[HMSCH == 1]),
            hh_kids_not_homeschool = length(HMSCH[HMSCH == 0]))

disp_ind_5_11 <- disp(fnl_dt_5_11_summary$RRACE,fnl_dt_5_11_summary$hh_kids_homeschool,fnl_dt_5_11_summary$hh_kids)

# Rank them from highest to lowest
disp_ind_5_11$Rank <- rank(-disp_ind_5_11$disproportionality)

kable(disp_ind_5_11[order(disp_ind_5_11$Rank, decreasing = FALSE), ],
      caption = "**Table 4: Disproportionality Indices for kids Homeschooling of age 5-11 yrs old by Race**")
```
******
```{r message=FALSE}
# Disproportionality Indices for kids Homeschooling of age 5-11 yrs old by Education Attainment
fnl_dt_5_11_summary <- fnl_dt_5_11%>%
  group_by(EEDUC)%>%
  summarise(hh_kids = n(),
            hh_kids_homeschool = length(HMSCH[HMSCH == 1]),
            hh_kids_not_homeschool = length(HMSCH[HMSCH == 0]))

disp_ind_5_11 <- disp(fnl_dt_5_11_summary$EEDUC,fnl_dt_5_11_summary$hh_kids_homeschool,fnl_dt_5_11_summary$hh_kids)

# Rank them from highest to lowest
disp_ind_5_11$Rank <- rank(-disp_ind_5_11$disproportionality)

kable(disp_ind_5_11[order(disp_ind_5_11$Rank, decreasing = FALSE), ],
      caption = "**Table 5: Disproportionality Indices for kids Homeschooling of age 5-11 yrs old by Education Attainment**")
```

### **STEP 5: MULTIVARIATE REGRESSION & ODDS RATIO**

Conduct a regression to test the hypothesis that the association of the dependent variable with independent variables is statistically significant. Also,evaluate the odds of being Homeschooling for above stated demographic factors.

**NOTE: ** For the purposes of this analysis, I'll use the reference categories of 

-   "White" for race 
-   "Well-Educated" for educational attainment
-   "Vaccinated" for Vaccination Status

```{r}
mdl_0_5_11 <- glm(HMSCH ~1, data = fnl_dt_5_11)
mdl_multi_5_11 <- glm(HMSCH ~ KIDDOSESRV_5_11Y+
                   relevel(RRACE,1)+
                   relevel(EEDUC,2),
                 data = fnl_dt_5_11,family = "binomial")
summary(mdl_multi_5_11)
exp(coef(mdl_multi_5_11))
lrtest(mdl_0_5_11,mdl_multi_5_11)
```

# **FOR KIDS AGE GROUP 12-17 YEARS OLD**
### **STEP 3: DESCRIPTIVE STATISTICS**

Proportion of kids Homeschooling based on the demographic factors.

```{r message=FALSE,fig.width=10, fig.height=5}
# For Kids Age Group :12-17 Yrs Old
# Plot Percent of Kids Homeschooling Vs Vaccination Status
plot1 <- fnl_dt_12_17%>%
  group_by(KIDDOSESRV_12_17Y,HMSCH)%>%
  summarise(count = n())%>%
  mutate(pct = (count/sum(count))*100)%>%
  filter(HMSCH == 1)%>%
  ggplot(aes(KIDDOSESRV_12_17Y, pct))+
  geom_bar(stat = "identity", position = 'dodge', fill="steelblue")+
  scale_y_continuous(breaks = seq(0,20,5))+
  labs(y = "% of Kids in Homeschooling", x = "Vaccination Status")+
  geom_text(aes(label = formatC(pct, digits = 1, format = "f")),
            colour = "white", size = 4, vjust = 1.5)+
  theme_light()

# Plot Percent of Kids Homeschooling Vs Race
plot2 <- fnl_dt_12_17%>%
  group_by(RRACE,HMSCH)%>%
  summarise(count = n())%>%
  mutate(pct = (count/sum(count))*100)%>%
  filter(HMSCH == 1)%>%
  ggplot(aes(RRACE, pct))+
  geom_bar(stat = "identity", position = 'dodge', fill="steelblue")+
  scale_y_continuous(breaks = seq(0,20,5))+
  labs(y = "% of Kids in Homeschooling", x = "Race")+
  geom_text(aes(label = formatC(pct, digits = 1, format = "f")),
            colour = "white", size = 4, vjust = 1.5)+
  theme_light()

# Plot Percent of Kids Homeschooling Vs Education Attainment
plot3 <- fnl_dt_12_17%>%
  group_by(EEDUC,HMSCH)%>%
  summarise(count = n())%>%
  mutate(pct = (count/sum(count))*100)%>%
  filter(HMSCH == 1)%>%
  ggplot(aes(EEDUC, pct))+
  geom_bar(stat = "identity", position = 'dodge', fill="steelblue")+
  scale_y_continuous(breaks = seq(0,20,5))+
  labs(y = "% of Kids in Homeschooling", x = "Education Attainment")+
  geom_text(aes(label = formatC(pct, digits = 1, format = "f")),
            colour = "white", size = 4, vjust = 1.5)+
  theme_light()

# Combining all plot into one
combined_plot_12_17 <- ggarrange(plot1, plot2, plot3, nrow = 1, ncol = 3,
                                common.legend = TRUE,legend="bottom")
annotate_figure(combined_plot_12_17,
                top = text_grob("Percent of Kids Homeschool by: Kids Vaccination Status,Race,Education Attainment\nAge Group: 12_17Yrs Old\n",
                                color = "black", face = "bold", size = 14))+
  theme_light()
```

### **STEP 4: DISPROPORTIONALITY INDICES**

Determine the disproportional indices for the % of children who are homeschooled based on the demographic characteristics 

```{r message=FALSE}
# Disproportionality Indices for kids Homeschooling of age 12-17 yrs old by Vaccination Status
fnl_dt_12_17_summary <- fnl_dt_12_17%>%
  group_by(KIDDOSESRV_12_17Y)%>%
  summarise(hh_kids = n(),
            hh_kids_homeschool = length(HMSCH[HMSCH == 1]),
            hh_kids_not_homeschool = length(HMSCH[HMSCH == 0]))

disp_ind_12_17 <- disp(fnl_dt_12_17_summary$KIDDOSESRV_12_17Y,fnl_dt_12_17_summary$hh_kids_homeschool,fnl_dt_12_17_summary$hh_kids)

# Rank them from highest to lowest
disp_ind_12_17$Rank <- rank(-disp_ind_12_17$disproportionality)

kable(disp_ind_12_17[order(disp_ind_12_17$Rank, decreasing = FALSE), ],
      caption = "**Table 6: Disproportionality Indices for kids Homeschooling of age 12-17 yrs old by Vaccination Status**")
```
******
```{r message=FALSE}
# Disproportionality Indices for kids Homeschooling of age 12-17 yrs old by Race
fnl_dt_12_17_summary <- fnl_dt_12_17%>%
  group_by(RRACE)%>%
  summarise(hh_kids = n(),
            hh_kids_homeschool = length(HMSCH[HMSCH == 1]),
            hh_kids_not_homeschool = length(HMSCH[HMSCH == 0]))

disp_ind_12_17 <- disp(fnl_dt_12_17_summary$RRACE,fnl_dt_12_17_summary$hh_kids_homeschool,fnl_dt_12_17_summary$hh_kids)

# Rank them from highest to lowest
disp_ind_12_17$Rank <- rank(-disp_ind_12_17$disproportionality)

kable(disp_ind_12_17[order(disp_ind_12_17$Rank, decreasing = FALSE), ],
      caption = "**Table 7: Disproportionality Indices for kids Homeschooling of age 12-17 yrs old by Race**")
```
******
```{r message=FALSE}
# Disproportionality Indices for kids Homeschooling of age 12-17 yrs old by Education Attainment
fnl_dt_12_17_summary <- fnl_dt_12_17%>%
  group_by(EEDUC)%>%
  summarise(hh_kids = n(),
            hh_kids_homeschool = length(HMSCH[HMSCH == 1]),
            hh_kids_not_homeschool = length(HMSCH[HMSCH == 0]))

disp_ind_12_17 <- disp(fnl_dt_12_17_summary$EEDUC,fnl_dt_12_17_summary$hh_kids_homeschool,fnl_dt_12_17_summary$hh_kids)

# Rank them from highest to lowest
disp_ind_12_17$Rank <- rank(-disp_ind_12_17$disproportionality)

kable(disp_ind_12_17[order(disp_ind_12_17$Rank, decreasing = FALSE), ],
      caption = "**Table 8: Disproportionality Indices for kids Homeschooling of age 12-17 yrs old by Education Attainment**")
```

### **STEP 5: MULTIVARIATE REGRESSION & ODDS RATIO**

Conduct a regression to test the hypothesis that the association of the dependent variable with independent variables is statistically significant.Also,evaluate the odds of being Homeschooling for above stated demographic factors.

**NOTE: ** For the purposes of this analysis, I'll use the reference categories of 

-   "White" for race 
-   "Well-Educated" for educational attainment
-   "Vaccinated" for Vaccination Status

```{r message=FALSE}
mdl_0_12_17 <- glm(HMSCH ~1, data = fnl_dt_12_17)
mdl_multi_12_17 <- glm(HMSCH ~ KIDDOSESRV_12_17Y+
                   relevel(RRACE,1)+
                   relevel(EEDUC,2),
                 data = fnl_dt_12_17,family = "binomial")
summary(mdl_multi_12_17)
exp(coef(mdl_multi_12_17))
lrtest(mdl_0_12_17,mdl_multi_12_17)
```
